---
title: Time Series Analysis of Seasonal Weather Patterns in the Four Corners Region of the United States
author: "Ben Karabinus"
date: "8/12/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
if(!require(DataExplorer)){install.packages("DataExplorer")}
library(DataExplorer)
if(!require(tidyverse)){install.packages("tidyverse")}
library(tidyverse)
if(!require(forecast)){install.packages("forecast")}
library(forecast)
if(!require(TSstudio)){install.packages("TSstudio")}
library(TSstudio)
if(!require(lubridate)){install.packages("lubridate")}
library(lubridate)
if(!require(plotly)){install.packages("plotly")}
library(plotly)
if(!require(stats)){install.packages("stats")}
library(stats)
if(!require(tseries)){install.packages("tseries")}
library(tseries)
```

## Problem Statement

In the past decades there has been a noticeable increase in the volatility of weather patterns in the Western United States. In particular, the four corners region (Arizona, Colorado, New Mexico and Utah) have experienced prolonged drought, monsoon rainfall, and a noticeable increase in the intensity and duration of heat waves. To better understand seasonal weather patterns in the region this report focuses on a time-series analysis of mean monthly temperature, total monthly precipitation and total monthly snowfall in the period between January 1, 1970 and December 31, 2019. Time series analysis was conducted using a Seasonal Autregressive Integrated Moving Average (SARIMA) model. The primary research questions addressed are as follows:

* Is there a noticeable trend in monthly mean temperature, total monthly precipitation, or total monthly snowfall?
* Is the SARIMA model an effective means of visualizng and predicting mean monthly temperature, total monthly precipitation and total monthly snowfall in the four corners region of the United States.

## Data Collection

Data for the analysis was sourced from the GLobal Historical Climatology Network (GHCN). The GHCN is an integrated database of daily climate summaries from land surface weather stations across the globe. GHCN data contain records from more than 100,000 stations in 180 countries and territories. All records are subject to a common suite of quality assurance reviews to miantain accuracy through uniformity of reporting stndards. All GHCN records include but are not limited too the following daily variables: 

* Daily maximum and Daily minimum temperatue 
* Total daily precipitation
* Total daily snowfall

### US Collection of GHCN Data

Daily weather records from the United States are compiled from a dozen separate datasets archived at the National Centers of Environmental Information (NCEI), a branch of the National Oceanic and Atmospheric Administration (NOAA). NCEI is responsible for hosting and providing access to one of the most significant stores of environmental data in existence with over 37 petabytes of oceanic, atmospheric and geophyscical data (National Centers for Environmental Information, 2019).

### Variables Included in the Model

* Average daily temperature (degrees farenheight)
* Total Daily precipitation (inches)
* Total Daily Snowfall (inches)

For more detailed information on each of the variables listed above please see (Appendix A)

## Data Set Features and Data Preparation

### Data Set Summary 

The data used in this report were observed at weather stations in four major cities located in the southwestern United States, specifically (Phoenix, Arizona), (Denver, Colorado), (Albequerque,New Mexico) and (Salt Lake City, Utah). Data was collected via land-based weather stations at major airports in each city with the exception being data collected in Denver, Colorado. Weather observations for Denver were collected using a land-based weather station located in the Sand Creek Open Space in the Denver Central Park Region in the Northwest corner of the city. Data for each city where downloaded as seperate comma seperated value (csv) files. Exploration of key features of the data sets, data cleansing and initial visualization were carried out using the "DataExplorer" library available in R.

### Data Discovery and Cleansing Process 

Each of the four data sets was examined prior to conducting time series analysis. Initial analysis focused on locating and replacing missing values, transforming data to appropriate types for anlysis and agreggating daily measurements. 

```{r echo=FALSE}
# Load Datasets
# Albuquerque
ABQ <- read.csv("ABQ.csv", header = T)
# Denver
DEN <- read.csv("DEN.csv", header = T)
# Phoenix
PHX <- read.csv("PHX.csv", header = T)
# Salt Lake City 
SLC <- read.csv("SLC.csv", header = T)
```

#### Summary statistics for each data set 

```{r echo=TRUE}
# Albuquerque
summary(ABQ)
```


```{r echo=TRUE}
# Denver 
summary(DEN)
```


```{r echo=TRUE}
# Phoenix
summary(PHX)
```

```{r echo=TRUE}
# Salt Lake City 
summary(SLC)
```

#### Structure of Each Data Set (Number of Observations and Variable Type)

```{r echo=FALSE}
# Albuquerque
str(ABQ)
```

```{r echo=FALSE}
# Denver
str(DEN)
```

```{r echo=FALSE}
# Phoenix
str(PHX)
```

```{r echo=FALSE}
# Salt Lake City 
str(SLC)
```

#### Visualize Missing Values From Each Dataset

```{r echo=TRUE}
# Albuquerque
plot_missing(ABQ, title = "Albuquerque",  group=c("No Missing Values"=0, 
                          "PCT Missing Values"= 1))
```

```{r echo=TRUE}
# Denver
plot_missing(DEN, title = "Denver",  group=c("No Missing Values"=0, 
                          "PCT Missing Values"= 1))
```

```{r echo=TRUE}
# Phoenix
plot_missing(PHX, title = "Phoenix", group=c("No Missing Values"=0,
                          "PCT Missing Values"= 1))
```

```{r echo=TRUE}
# Salt Lake City 
plot_missing(SLC, title = "Salt Lake City ", group=c("No Missing Values"=0, 
                                                "PCT Missing Values"= 1))
```

#### Replace Misssing Values 

The above graphs show that each of the data sets are missing observations for one or more values. To avoid the effects of outliers, missing values for TMIN, TMAX, PRCP and SNOW will be replaced with the median value for each of the respective variables. According GHCN Daily TAVG is calculated by taking the mean value for daily observations of TMAX and TMIN and rounding up to the nearest integer, That is daily $TAVG=ceiling(\frac{TMAX+TMIN}{2})$. Therefore, missing values for TAVG will be replaced using this formula. For readability only the replacement of missing values from the "ABQ" data set is detailed below. Please see the .rmd version of this report for a more in depth eplanation of the replacement process.

```{r echo=TRUE}
# replace missing values in ABQ
# replace NA's in the SNOW column with the median snowfall 
ABQ$SNOW[is.na(ABQ$SNOW)] <- median(ABQ$SNOW, na.rm = T)
# replace the missing values for TAVG
ABQTEMP <- data.frame(ABQ$TMAX, ABQ$TMIN)
ind <- which(is.na(ABQ), arr.ind=TRUE)
# all temperatures are given as whole numbers in dataset so round the mean
ABQ[ind] <- ceiling(rowMeans(ABQTEMP, na.rm=TRUE)[ind[,1]])
#check to ensure replacement 
profile_missing(ABQ)
```

```{r echo=FALSE}
# replace missing values in DEN
# remove NA's for precipitation 
DEN$PRCP[is.na(DEN$PRCP)] <- median(DEN$PRCP, na.rm = TRUE)
# remove the NA's for SNOW
DEN$SNOW[is.na(DEN$SNOW)] <- median(DEN$SNOW, na.rm = TRUE)
# remove the NA's for TMAX
DEN$TMAX[is.na(DEN$TMAX)] <- median(DEN$TMAX, na.rm = TRUE)
# remove the NA's for TMIN
DEN$TMIN[is.na(DEN$TMIN)] <- median(DEN$TMIN, na.rm = TRUE)
# replace the missing values for TAVG
DENTEMP <- data.frame(DEN$TMAX, DEN$TMIN)
ind <- which(is.na(DEN), arr.ind=TRUE)
# all temperatures are given as whole numbers in dataset so round the mean
DEN[ind] <- ceiling(rowMeans(DENTEMP, na.rm=TRUE)[ind[,1]])
#check to ensure replacement 
#profile_missing(DEN)
```

```{r echo=FALSE}
# replace missing values in PHX
# remove the NA's for SNOW
PHX$SNOW[is.na(PHX$SNOW)] <- median(PHX$SNOW,na.rm = TRUE)
# replace the missing values for TAVG
PHXTEMP <- data.frame(PHX$TMAX, PHX$TMIN)
ind <- which(is.na(PHX), arr.ind=TRUE)
# all temperatures are given as whole numbers in dataset so round the mean
PHX[ind] <- ceiling(rowMeans(PHXTEMP, na.rm=TRUE)[ind[,1]])
#check to ensure replacement 
#profile_missing(PHX)
```

```{r echo=FALSE}
# replace missing values in SLC 
# remove the NA's for SNOW
SLC$SNOW[is.na(SLC$SNOW)] <- median(SLC$SNOW,na.rm = TRUE)
# replace the missing values for TAVG
SLCTEMP <- data.frame(SLC$TMAX, SLC$TMIN)
ind <- which(is.na(SLC), arr.ind=TRUE)
SLC[ind] <- ceiling(rowMeans(SLCTEMP, na.rm=TRUE)[ind[,1]])
#check to ensure replacement 
#profile_missing(SLC)
```

#### Preparing Data for Time Series Analysis

```{r echo=TRUE}
# transform date column from factor to date data type in each data set
ABQ <- transform(ABQ, DATE = as.Date(DATE))
DEN <- transform(DEN, DATE = as.Date(DATE))
PHX <- transform(PHX, DATE = as.Date(DATE))
SLC <- transform(SLC, DATE = as.Date(DATE))
#verify changes 
ABQ_DELTA <- sapply(ABQ, class)
DEN_DELTA <- sapply(DEN, class)
PHX_DELTA <- sapply(PHX, class)
SLC_DELTA <- sapply(SLC, class)
```

```{r echo=TRUE}
ABQ_DELTA
```

```{r echo=FALSE}
#DEN_DELTA
```

```{r echo=FALSE}
#PHX_DELTA
```

```{r echo=FALSE}
#SLC_DELTA
```

#### Aggregate Data for Monthly Analysis 

```{r echo=TRUE}
# Albuquerque
# create column month year to aggregate data for time series 
ABQ$MONTH_YEAR <- floor_date(ABQ$DATE,"month")
# create aggregated dataset using MONTH_YEAR COLUMN
ABQ_AGG <- ABQ %>%
  group_by(MONTH_YEAR)%>%
  dplyr::summarize(Monthly_Avg_Temp = mean(TAVG),
  Monthly_Rainfall=sum(PRCP), Monthly_Snowfall=sum(SNOW)) %>%
  as.data.frame()
head(ABQ_AGG)
```



```{r echo=FALSE}
# Denver 
# create column month year to aggregate data for time series 
DEN$MONTH_YEAR <- floor_date(DEN$DATE,"month")
# create aggregated dataset using MONTH_YEAR COLUMN
# This aggeregate will be used to create the time series 
DEN_AGG <- DEN %>%
  group_by(MONTH_YEAR)%>%
  dplyr::summarize(Monthly_Avg_Temp = mean(TAVG),
  Monthly_Rainfall=sum(PRCP), Monthly_Snowfall=sum(SNOW)) %>%
  as.data.frame()
```


```{r echo=FALSE}
# Phoenix 
# create column month year to aggregate data for time series 
PHX$MONTH_YEAR <- floor_date(PHX$DATE,"month")
# create aggregated dataset using MONTH_YEAR COLUMN
PHX_AGG <- PHX %>%
  group_by(MONTH_YEAR)%>%
  dplyr::summarize(Monthly_Avg_Temp = mean(TAVG),
  Monthly_Rainfall=sum(PRCP), Monthly_Snowfall=sum(SNOW)) %>%
  as.data.frame()
```


```{r echo=FALSE}
# Salt Lake City 
# create column month year to aggregate data for time series 
SLC$MONTH_YEAR <- floor_date(SLC$DATE,"month")
# create aggregated dataset using MONTH_YEAR COLUMN
SLC_AGG <- SLC %>%
  group_by(MONTH_YEAR)%>%
  dplyr::summarize(Monthly_Avg_Temp = mean(TAVG),
  Monthly_Rainfall=sum(PRCP), Monthly_Snowfall=sum(SNOW)) %>%
  as.data.frame()
```

### Analysis Using Seasonal Autoregressive Integrated Moving Average SARIMA Model

#### What is a SARIMA Model?

A SARIMA model is the seasonal flavor of the autoregressive integrated moving average (ARIMA) model. ARIMA and SARIMA models are a statistical technique used to analyze times series data and in certain cases predict future values. The goal of autoregressive models is to predict future values of the target variable by regressing against past observations of the variable. That is lagged values of the target variable are used as independent variables "x's" in the regression equation (Yiu, 2020). Seasonal autoregressive models are used to perform similar analysis while accounting for any seasonality obesrved in the time series being analyzed. ARIMA and SARIMA models also incorporate differencing to correct for data that is not stationary and a moving average component that includes errors in previous predictions as a parameter in the model. To better understand the method a brief summary of each component is given below.


#### Autoregressive (p)

Considering an observed value "Y" to be a linear function of it's past observations a simple representation of the regression equation is as follows.

$$Y=B0+B1*Y_{lag1}+B2*Y_{lag2}......Bn*Y_{lagn}$$

By tweaking this equation we can attemmt to predict future values of Y.

$$Y_{forward1}=B0+B1*Y+B2*Y_{lag1}......Bn*Y_{lag(n-1)}$$

The above equation represents whats commonly referred to as an AR(1) model. An autoregressive model using a single lag to predict future values of Y (Yiu, 2020). 

#### Integrated (d)

When conducting times series analysis an important step is ensuring that time series data is stationary. Stationary time series data has a more stable mean and variance. That is the mean and variance of the data are consistent over time leading to a more robust model. Performing one or more differencing steps on time series data helps to ensure stationarity. In ARIMA and SARIMA modeling this step can be performed as follows.

$$Y_{forward1}=B0+B1*(Y-Y_{lag1})+B2*(Y_{lag1}-Y_{lag2})......Bn*(Y_{lag(n-1)}-Y_{lag(n-2)})$$

#### Moving Average (q)

Moving avaerage(MA) is similar to the auto regresive step in that it uses past values (lags) to predict future outcomes. The difference is that the MA method uses previous error terms in the regression model as parameters for predicting future values. Error terms or "E" represent random deviations between values of the target variable fitted by the model and actual observations at each lag. 

$$Y=B0+B1*E_{lag1}+B2*E_{lag2}......Bn*E_{lagn}$$

For a simple moving average model the above equation can then be modified to include the average of "Y." Thus errors at each lag push the average of Y in the positive or negative direction adjusting the moving average of Y. 

$$Y=mean(Y)+B1*E_{lag1}$$

#### Seasonal (PDQ)[Seasonality]

When constructing an ARIMA model it's important to note the presence of cyclical or predictable patterns in the data being analyzed. Many times these patterns represent seasonality. If you wanted to model the sale of baseball tickets on a monthly basis you would notice a sharp decrease to zero tickets sold between the months of February and March. That's becuse this is the offseason for baseball. The offseason begins and ends at roughly the same time every year. So when using an ARIMA model we should account for this seasonality by adding seasonal parameters to the model. Using the auto.arima function in R approximates the seasonal parameters of an ARIMA model helping to account for seasonality and improve model performance. An abstraction of a SARIMA model and it's parameters as output by the auto.arima function is given below.

$$ARIMA(p,d,q)(P,D,Q)[S]$$

### Preparing to Model the Data Using SARIMA

The first step when using the auto.arima function in R is to tranform the data being analyzed into a "time series object.Time series objects will be created for mean mothly temperature, total monthly precipation (inches) and total monthly snowfall (inches) for each of the four locations being analyzed.

```{r echo=TRUE}
# create time series objects for mean monthly temperature 
ABQ_TS_TEMP <- ts(ABQ_AGG[, 2], start= c(1970,1), end= c(2019,12), frequency = 12)
DEN_TS_TEMP <- ts(DEN_AGG[, 2], start= c(1970,1), end= c(2019,12), frequency = 12)
PHX_TS_TEMP <- ts(PHX_AGG[, 2], start= c(1970,1), end= c(2019,12), frequency = 12)
SLC_TS_TEMP <- ts(SLC_AGG[, 2], start= c(1970,1), end= c(2019,12), frequency = 12)
# create time series objects for total monthly precipitation 
ABQ_TS_PRCP <- ts(ABQ_AGG[, 3], start=c(1970, 1), end=c(2019,12), frequency=12)
DEN_TS_PRCP <- ts(DEN_AGG[, 3], start=c(1970, 1), end=c(2019,12), frequency=12)
PHX_TS_PRCP <- ts(PHX_AGG[, 3], start=c(1970, 1), end=c(2019,12), frequency=12)
SLC_TS_PRCP <- ts(SLC_AGG[, 3], start=c(1970, 1), end=c(2019,12), frequency=12)
# create time series objects for total monthly snowfall 
ABQ_TS_SNOW <- ts(ABQ_AGG[, 4], start=c(1970,1), end=c(2019,12), frequency=12)
DEN_TS_SNOW <- ts(DEN_AGG[, 4], start=c(1970,1), end=c(2019,12), frequency=12)
PHX_TS_SNOW <- ts(PHX_AGG[, 4], start=c(1970,1), end=c(2019,12), frequency=12)
SLC_TS_SNOW <- ts(SLC_AGG[, 4], start=c(1970,1), end=c(2019,12), frequency=12)
```


#### Test That the Time Series Data Are Stationary

Time series may be considered stationary if they have no significant trend, constant variance over time and a constant autocorrelation structure over time (Zach, 2021). One mehtod of testing whether times series data are stationary in R is the to use the adf.test function from the tseries library. The adf.test function performs an augmented Dickey-Fuller test on the specified  time series object. The augmented Dickey-Fuller Test gives the following null hypothesis, alternate hypothesis and confidence level.

$$H_o: \space The \space time \space series \space is \space non-stationary.$$

$$H_A: \space the \space time \space series \space is \space stationary$$

When applied to each of the time series above the output of the adf.test function showed the null hypothesis should be rejected. Instead the alternate hypothesis that the data are stationary should be accepted. The adf.test function as applied to the "ABQ_TS_TEMP" time series object is given as an example below. For a more detailed explanation of the process please see the .rmd version of this report.

```{r echo=TRUE}
# perform the ADF test for ABQ
adf.test(ABQ_TS_TEMP)
```

```{r echo=FALSE}
# Perform the ADF test for remaining time series objects
#adf.test(ABQ_TS_PRCP)
#adf.test(ABQ_TS_SNOW)
#adf.test(DEN_TS_TEMP)
#adf.test(DEN_TS_PRCP)
#adf.test(DEN_TS_SNOW)
#adf.test(PHX_TS_TEMP)
#adf.test(PHX_TS_PRCP)
#adf.test(PHX_TS_SNOW)
#adf.test(SLC_TS_TEMP)
#adf.test(SLC_TS_PRCP)
#adf.test(SLC_TS_SNOW)
```

#### Visualize the Data

An important step in time series analysis is to visually inspect the data. Time series visualization allows for further assessment of whether the data are stationary, decomposition of time series attributes and ispection for anomalies in the data.

* High-level visual of the time series objects created for measurements from ALbuquerque

```{r echo=FALSE}
# plot the time series data for Albuquerque (birds eye view)
par(mfrow=c(3,1))
plot(ABQ_TS_TEMP)
plot(ABQ_TS_PRCP)
plot(ABQ_TS_SNOW)
```

* Shortened time-span to allow for closer look at the behavior of each variable

```{r echo=FALSE}
# shorten the time span and customize the plot (close up view)
ABQ_TS_TEMP_SHORT <- ts(ABQ_AGG[, 2], start= c(2010,1), end= c(2019,12), frequency = 12)
ABQ_TEMP_PLT <- ts_plot(ABQ_TS_TEMP_SHORT, title = "Mean Monthly Temperature Albuqurque, NM",                           Xtitle = "Time",
                        Ytitle="Temperature (F)",
                        color="#ff0000",
                        Xgrid=TRUE,
                        Ygrid=TRUE,) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="White",
                    zerolinecolor="White",
                    gridcolor="White"))
ABQ_TS_PRCP_SHORT <- ts(ABQ_AGG[, 3], start= c(2010,1), end= c(2019,12), frequency = 12)
ABQ_PRCP_PLT <- ts_plot(ABQ_TS_PRCP_SHORT, title = "Monthly Rainfall Albuquerque, NM", Xtitle = "Time",
                     Ytitle="Total Montlhy Rainfall (inches)",
                     color="#07d870",
                     Xgrid=TRUE,
                     Ygrid=TRUE) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="White",
                    zerolinecolor="White",
                    gridcolor="White"))
ABQ_TS_SNOW_SHORT <- ts(ABQ_AGG[, 4], start= c(2010,1), end= c(2019,12), frequency = 12)
ABQ_SNOW_PLT <- ts_plot(ABQ_TS_SNOW_SHORT, title = "Monthly Snowfall Albuquerque, NM", 
                        Xtitle ="Time",
                        Ytitle="Total Montlhy Snowfall (inches)",
                        color="#4982ff",
                        Xgrid=TRUE,
                        Ygrid=TRUE) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="white",
                    zeroline=TRUE,
                    zerolinecolor="white",
                    gridcolor="white"))

par(mfrow=c(3,1))
ABQ_SNOW_PLT
ABQ_PRCP_PLT
ABQ_TEMP_PLT
```

* Decompostition of time series from the city of Albuquerque shows that the series for mean monthly temperature, total  monthly precipitation and total monthly snowfall each exhibit a strong seasonal component.Total monthly prcipitation and total monthly snowfall both show much more variance in observed values over time. It's worth noting that there seems to be an increase in mean monthly temperature around the year 2016.

```{r echo=FALSE}
ts_decompose(ABQ_TS_TEMP_SHORT)
ts_decompose(ABQ_TS_PRCP_SHORT)
ts_decompose(ABQ_TS_SNOW_SHORT)
```


* High-level visual of the time series objects created for measurements from Denver

```{r echo=FALSE}
# plot the time series data for Denver (birds eye view)
par(mfrow=c(3,1))
plot(DEN_TS_TEMP)
plot(DEN_TS_PRCP)
plot(DEN_TS_SNOW)
```


* Shortened time-span to allow for closer look at the behavior of each variable


```{r echo=FALSE}
# shorten the time span and customize the plot (close up view)
DEN_TS_TEMP_SHORT <- ts(DEN_AGG[, 2], start= c(2010,1), end= c(2019,12), frequency = 12)
DEN_TEMP_PLT <- ts_plot(DEN_TS_TEMP_SHORT, title = "Mean Monthly Temperature Denver, CO",                               Xtitle = "Time",
                        Ytitle="Temperature (F)",
                        color="#ff0000",
                        Xgrid=TRUE,
                        Ygrid=TRUE,) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="White",
                    zerolinecolor="White",
                    gridcolor="White"))
DEN_TS_PRCP_SHORT <- ts(DEN_AGG[, 3], start= c(2010,1), end= c(2019,12), frequency = 12)
DEN_PRCP_PLT <- ts_plot(DEN_TS_PRCP_SHORT, title = "Monthly Rainfall Denver, CO",                                    Xtitle = "Time",
                     Ytitle="Total Montlhy Rainfall (inches)",
                     color="#07d870",
                     Xgrid=TRUE,
                     Ygrid=TRUE) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="White",
                    zerolinecolor="White",
                    gridcolor="White"))
DEN_TS_SNOW_SHORT <- ts(DEN_AGG[, 4], start= c(2010,1), end= c(2019,12), frequency = 12)
DEN_SNOW_PLT <- ts_plot(DEN_TS_SNOW_SHORT, title = "Monthly Snowfall Denver, CO", 
                        Xtitle ="Time",
                        Ytitle="Total Montlhy Snowfall (inches)",
                        color="#4982ff",
                        Xgrid=TRUE,
                        Ygrid=TRUE) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="white",
                    zeroline=TRUE,
                    zerolinecolor="white",
                    gridcolor="white"))

par(mfrow=c(3,1))
DEN_SNOW_PLT
DEN_PRCP_PLT
DEN_TEMP_PLT
```

* Decompostition of time series from the city of Denver shows that the series for mean monthly temperature, total  monthly precipitation and total monthly snowfall each exhibit a strong seasonal component similar to the time series data from Albuquerque. Examining the plots of the decomposed series below shows that the trend in each series seems exhibit a somewhat cyclical pattern. Eluding to the fact that there may be a seasonal pattern occuring on multi-year scale sucha as, the effects of the El Nino-Southern Oscilation cycle.


```{r echo=TRUE}
par(mfrow=c(1,3))
ts_decompose(DEN_TS_TEMP_SHORT)
ts_decompose(DEN_TS_PRCP_SHORT)
ts_decompose(DEN_TS_SNOW_SHORT)
```


* High-level visual of the time series objects created for measurements from Phoenix 


```{r echo=FALSE}
# plot the time series data for Phoenix (birds eye view)
par(mfrow=c(3,1))
plot(PHX_TS_TEMP)
plot(PHX_TS_PRCP)
plot(PHX_TS_SNOW)
```



```{r echo=FALSE}
# shorten the time span and customize the plot (close up view)
PHX_TS_TEMP_SHORT <- ts(PHX_AGG[, 2], start= c(2010,1), end= c(2019,12), frequency = 12)
PHX_TEMP_PLT <- ts_plot(PHX_TS_TEMP_SHORT, title = "Mean Monthly Temperature Phoenix, AZ",                               Xtitle = "Time",
                        Ytitle="Temperature (F)",
                        color="#ff0000",
                        Xgrid=TRUE,
                        Ygrid=TRUE,) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="White",
                    zerolinecolor="White",
                    gridcolor="White"))
PHX_TS_PRCP_SHORT <- ts(PHX_AGG[, 3], start= c(2010,1), end= c(2019,12), frequency = 12)
PHX_PRCP_PLT <- ts_plot(PHX_TS_PRCP_SHORT, title = "Monthly Rainfall Phoenix, AZ",                                   Xtitle = "Time",
                     Ytitle="Total Montlhy Rainfall (inches)",
                     color="#07d870",
                     Xgrid=TRUE,
                     Ygrid=TRUE) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="White",
                    zerolinecolor="White",
                    gridcolor="White"))
PHX_TS_SNOW_SHORT <- ts(PHX_AGG[, 4], start= c(2010,1), end= c(2019,12), frequency = 12)
PHX_SNOW_PLT <- ts_plot(PHX_TS_SNOW_SHORT, title = "Monthly Snowfall Phoenix, AZ", 
                        Xtitle ="Time",
                        Ytitle="Total Montlhy Snowfall (inches)",
                        color="#4982ff",
                        Xgrid=TRUE,
                        Ygrid=TRUE) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="white",
                    zeroline=TRUE,
                    zerolinecolor="white",
                    gridcolor="white"))

par(mfrow=c(3,1))
PHX_SNOW_PLT
PHX_PRCP_PLT
PHX_TEMP_PLT
```

 * Decompostition of time series from the city of Phoenix shows that the series for mean monthly temperature, total  monthly precipitation and total monthly snowfall each exhibit a strong seasonal component similar to the time series data from both Denver and Albuquerque. Examining the plots of the decomposed series below it's worth noting that there seems to be a slighty upward trend in both mean monthly temperature and total monthly precipitation ocurring around around the year 2016. The trend in monthly precipitation may be attributed to a multi-year seasonal pattern, but the rising trend in mean monthly temperature appears at a glance to be more of a departure from the trend of past observations. The decomposition of the series for monthly snowfall confirms that it rarely snows in Phoenix. The series will be omitted from further analysis in this report.

```{r echo=FALSE}
ts_decompose(PHX_TS_TEMP_SHORT)
ts_decompose(PHX_TS_PRCP_SHORT)
ts_decompose(PHX_TS_SNOW_SHORT)
```

* High-level visual of the time series objects created for measurements from Salt Lake City

```{r echo=FALSE}
# plot the time series data for Salt Lake City (birds eye view)
par(mfrow=c(3,1))
plot(SLC_TS_TEMP)
plot(SLC_TS_PRCP)
plot(SLC_TS_SNOW)
```



```{r echo=FALSE}
# shorten the time span and customize the plot (close up view)
SLC_TS_TEMP_SHORT <- ts(SLC_AGG[, 2], start= c(2010,1), end= c(2019,12), frequency = 12)
SLC_TEMP_PLT <- ts_plot(SLC_TS_TEMP_SHORT, title ="Mean Monthly Temperature Salt Lake City, UT",                         Xtitle = "Time",
                        Ytitle="Temperature (F)",
                        color="#ff0000",
                        Xgrid=TRUE,
                        Ygrid=TRUE,) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="White",
                    zerolinecolor="White",
                    gridcolor="White"))
SLC_TS_PRCP_SHORT <- ts(SLC_AGG[, 3], start= c(2010,1), end= c(2019,12), frequency = 12)
SLC_PRCP_PLT <- ts_plot(SLC_TS_PRCP_SHORT, title = "Monthly Rainfall Salt Lake City, UT",                            Xtitle = "Time",
                     Ytitle="Total Montlhy Rainfall (inches)",
                     color="#07d870",
                     Xgrid=TRUE,
                     Ygrid=TRUE) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="White",
                    zerolinecolor="White",
                    gridcolor="White"))
SLC_TS_SNOW_SHORT <- ts(SLC_AGG[, 4], start= c(2010,1), end= c(2019,12), frequency = 12)
SLC_SNOW_PLT <- ts_plot(SLC_TS_SNOW_SHORT, title = "Monthly Snowfall Salt Lake City, UT", 
                        Xtitle ="Time",
                        Ytitle="Total Montlhy Snowfall (inches)",
                        color="#4982ff",
                        Xgrid=TRUE,
                        Ygrid=TRUE) %>%
  layout(paper_bgcolor="black",
         plot_bgcolor="black",
         font=list(color="white"),
         yaxis=list(linecolor="white",
                    zerolinecolor="white",
                    gridcolor="white"),
         xaxis=list(linecolor="white",
                    zeroline=TRUE,
                    zerolinecolor="white",
                    gridcolor="white"))

par(mfrow=c(3,1))
SLC_SNOW_PLT
SLC_PRCP_PLT
SLC_TEMP_PLT
```

```{r echo=FALSE}
ts_decompose(SLC_TS_TEMP_SHORT)
ts_decompose(SLC_TS_PRCP_SHORT)
ts_decompose(SLC_TS_SNOW_SHORT)
```


#### Create the SARIMA MODEL for Each Time Series 

```{r cache=TRUE}
# SARIMA model(s) for Albuquerque
ABQ_TEMP_MODEL <- auto.arima(ABQ_TS_TEMP, stepwise = FALSE)
ABQ_PRCP_MODEL <- auto.arima(ABQ_TS_PRCP, stepwise = FALSE)
ABQ_SNOW_MODEL <- auto.arima(ABQ_TS_SNOW, stepwise = FALSE)
```

```{r echo=TRUE}
# summary of ABQ_TEMP_MODEL
ABQ_TEMP_MODEL
```

```{r echo=FALSE}
# forecast using the ABQ_TEMP_MODEL
ABQ_TEMP_CAST <- forecast(ABQ_TEMP_MODEL, h=60)
plt <- plot_forecast(ABQ_TEMP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Mean Monthly Temperature",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```


```{r echo=TRUE}
# summary of ABQ_PRCP_MODEL
ABQ_PRCP_MODEL
```

```{r echo=FALSE}
# forecast using the ABQ_PRCP_MODEL
ABQ_PRCP_CAST <- forecast(ABQ_PRCP_MODEL, h=60)
ABQ_PRCP_CAST
ABQ_PRCP_CAST$lower<-apply(ABQ_PRCP_CAST$lower, 2, function(x) ifelse(x < 0,0, x))
plt <- plot_forecast(ABQ_PRCP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Total Monthly Precipitation",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```



```{r echo=TRUE}
# summary of ABQ_SNOW_MODEL
ABQ_SNOW_MODEL
```

```{r echo=FALSE}
ABQ_SNOW_CAST <- forecast(ABQ_SNOW_MODEL, h=60)
ABQ_SNOW_CAST
ABQ_SNOW_CAST$lower<-apply(ABQ_PRCP_CAST$lower, 2, function(x) ifelse(x < 0,0, x))
plt <- plot_forecast(ABQ_SNOW_CAST, title = "Five Year Forecast", 
                      Ytitle = "Total Monthly Snowfall",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```


```{r cache=TRUE}
# SARIMA model(s) for Denver
DEN_TEMP_MODEL <- auto.arima(DEN_TS_TEMP, stepwise = FALSE)
DEN_PRCP_MODEL <- auto.arima(DEN_TS_PRCP, stepwise = FALSE)
DEN_SNOW_MODEL <- auto.arima(DEN_TS_SNOW, stepwise = FALSE)
```

```{r echo=TRUE}
# summary of DEN_TEMP_MODEL
DEN_TEMP_MODEL
```


```{r echo=FALSE}
# forecast using the DEN_TEMP_MODEL
DEN_TEMP_CAST <- forecast(ABQ_TEMP_MODEL, h=60)
DEN_TEMP_CAST
plt <- plot_forecast(DEN_TEMP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Mean Monthly Temperature",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```


```{r echo=TRUE}
# summary of DEN_PRCP_MODEL
DEN_PRCP_MODEL
```


```{r echo=FALSE}
# forecast using the DEN_PRCP_MODEL
DEN_PRCP_CAST <- forecast(ABQ_PRCP_MODEL, h=60)
DEN_PRCP_CAST
DEN_PRCP_CAST$lower<-apply(DEN_PRCP_CAST$lower, 2, function(x) ifelse(x < 0,0, x))
plt <- plot_forecast(ABQ_PRCP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Total Monthly Precipitation",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```

```{r echo=FALSE}
# summary of DEN_SNOW_MODEL
summary(DEN_SNOW_MODEL)
```

```{r echo=FALSE}
# forecast using the DEN_SNOW_MODEL
DEN_SNOW_CAST <- forecast(DEN_SNOW_MODEL, h=60)
DEN_SNOW_CAST
DEN_SNOW_CAST$lower<-apply(DEN_SNOW_CAST$lower, 2, function(x) ifelse(x < 0,0, x))
plt <- plot_forecast(ABQ_PRCP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Total Monthly Snowfall",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```


```{r cache=TRUE}
# SARIMA model(s) for Phoenix
PHX_TEMP_MODEL <- auto.arima(PHX_TS_TEMP, stepwise = FALSE)
PHX_PRCP_MODEL <- auto.arima(PHX_TS_PRCP, stepwise = FALSE)
PHX_SNOW_MODEL <- auto.arima(PHX_TS_SNOW, stepwise = FALSE)
```

```{r echo=TRUE}
summary(PHX_TEMP_MODEL)
```

```{r echo=FALSE}
# forecast using the PHX_TEMP_MODEL
PHX_TEMP_CAST <- forecast(PHX_TEMP_MODEL, h=60)
PHX_TEMP_CAST
plt <- plot_forecast(PHX_TEMP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Mean Monthly Temperature",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```


```{r echo=TRUE}
summary(PHX_PRCP_MODEL)
```

```{r echo=FALSE}
# forecast using the PHX_PRCP_MODEL
PHX_PRCP_CAST <- forecast(PHX_PRCP_MODEL, h=60)
PHX_PRCP_CAST
PHX_PRCP_CAST$lower<-apply(PHX_PRCP_CAST$lower, 2, function(x) ifelse(x < 0,0, x))
plt <- plot_forecast(PHX_PRCP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Total Montly Precipitation",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```


```{r cache=TRUE}
SLC_TEMP_MODEL <- auto.arima(SLC_TS_TEMP, stepwise = FALSE)
SLC_PRCP_MODEL <- auto.arima(SLC_TS_PRCP, stepwise = FALSE)
SLC_SNOW_MODEL <- auto.arima(SLC_TS_SNOW, stepwise = FALSE)
```


```{r echo=TRUE}
summary(SLC_TEMP_MODEL)
```

```{r echo=TRUE}
# forecast using the SLC_TEMP_MODEL
SLC_TEMP_CAST <- forecast(SLC_TEMP_MODEL, h=60)
SLC_TEMP_CAST
SLC_TEMP_CAST$lower<-apply(SLC_TEMP_CAST$lower, 2, function(x) ifelse(x < 0,0, x))
plt <- plot_forecast(SLC_TEMP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Mean Monthly Temperature",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```

```{r echo=TRUE}
summary(SLC_PRCP_MODEL)
```

```{r echo=FALSE}
# forecast using the SLC_PRCP_MODEL
SLC_PRCP_CAST <- forecast(SLC_PRCP_MODEL, h=60)
SLC_PRCP_CAST
SLC_PRCP_CAST$lower<-apply(SLC_PRCP_CAST$lower, 2, function(x) ifelse(x < 0,0, x))
plt <- plot_forecast(SLC_PRCP_CAST, title = "Five Year Forecast", 
                      Ytitle = "Total Montly Precipitation",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```

```{r echo=TRUE}
summary(SLC_SNOW_MODEL)
```

```{r echo=FALSE}
# forecast using the SLC_SNOW_MODEL
SLC_SNOW_CAST <- forecast(SLC_SNOW_MODEL, h=60)
SLC_SNOW_CAST
SLC_SNOW_CAST$lower<-apply(SLC_SNOW_CAST$lower, 2, function(x) ifelse(x < 0,0, x))
plt <- plot_forecast(SLC_SNOW_CAST, title = "Five Year Forecast", 
                      Ytitle = "Total Montly Snowfall",
                       Xtitle = "Time") %>% layout(
                       xaxis = list(range=c(2015, 2024))
)
plt
```


### Conclusion

